### 2023-01-21 paper/localpsf.tex
small edits to the tex file, cleanup, consistent use of basal sliding and no longer using basal friction and also adding the arXiv version of the HODLR paper as a reference instead of referencing the submitted HODLR paper

Recently there have been improvements in matrix-free H-matrix construction methods in which an operator is applied to structured random vectors, and the response of the operator to those random vectors is processed to construct an H-matrix approximation  \cite{LevittMartinsson22,LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}. 
These methods (which we do not use here) have been used to approximate Hessians in PDE constrained inverse problems \cite{AmbartsumyanEtAl20,HartlandEtAl22}. 
Although the required number of matrix-vector products grows nearly linearly in the discretized parameter dimension and linearly in the rank of the off-diagonal blocks, the required number of matrix-vector products is still large (e.g., hundreds to thousands).
%These methods are asymptotically scalable in the sense that the required number of matrix-vector products grows nearly linearly in the dimension of the discretized parameter and linearly in the rank of the off-diagonal blocks. But despite asymptotic scalability, from a practical perspective the required number of matrix-vector products is large (e.g., hundreds to thousands). 
%Improving the peeling process to reduce its cost is currently an area of active research \cite{PEELINGRESEARCH}.
%This is an area of active research, so more efficient and practical matrix-free H-matrix construction methods based on the peeling process may be discovered in the future.
%Matrix-free H-matrix construction methods are currently an area of active research (see, e.g., \cite{LevittMartinsson22}), so more efficient matrix-free H-matrix methods may be discovered in the future. 
In this paper, we also form an H-matrix Hessian approximation. However, to reduce the required number of operator actions, we first form a PSF approximation of the data misfit Hessian by exploiting locality and non-negative integral kernel properties, then form the H-matrix using classical H-matrix techniques. Not all data misfit Hessians satisfy the local non-negative integral kernel properties (for example, the wave inverse problem data misfit Hessian has a substantial amount of negative entries in its integral kernel). However, many data misfit Hessians of practical interest do satisfy these properties (either exactly or approximately), and our method is targeted at approximating these Hessians.

@@

%\textbf{Heat equation}
%\begin{itemize}
%	\item Hessian per-column error plots for 1, 5, and 15 batches
%	\item $H-P$ error vs number of batches for interior and whole domain
%	\item $P^{-1}H-I$: error vs rank for $P=R$ and $P=P_\text{pch}$ diffusion time, for 1, 5, and 15 batches
%	\item Krylov iterations to tols 1e-1 and 1e-6 vs diffusion parameter
%	\item GLR vs PCH vs diffusion time
%	\item Krylov method convergence plot: $R$ vs $P_\text{pch}$ vs no preconditioner
%	\item Krylov iterations to tols 1e-6 vs mesh size $h$, PCH vs R vs no preconditioner (hold)
%	\item true parameter (H)
%	\item deterministic reconstruction (H)
%	\item noisy measurements (1\%, 5\%, 10\% noise). $u|_\text{top}$ (H)
%	\item recovered state at top (H)
%	\item mesh scalability of PCH (CC) (H)
%	\item put plot data into data directory
%	\item save function .pvd files in paraview  directory
%\end{itemize}
%
%\begin{itemize}
%	\item GLR vs PCH number of obs (Computational cost in PDE solves) (S)
%	\item GLR vs PCH aspect ratio (CC) (S)
%	\item turn on preconditioner after number of krylov iterations exceeds 15 in an iteration. Rebuild "as needed". (S)
%\end{itemize}
%
%\begin{itemize}
%	\item Stokes Krylov convergence (10 solves for nonlinear forward + 6 hessian matvecs for ellipsoid estimation + 10 matvecs for impulse responses + 30 matvecs for PCH6 CG solve + 100 matvecs for Reg CG solve)
%	\item Stokes error vs num batches (small number of batches; 1, 3, 6, 9 batches?) (50 matvecs for randomized error estimation +)
%	\item Velocity field
%	\item Basal friction field
%	\item Picture of impulse responses
%	\item Mesh negative numbers picture
%\end{itemize}
%
%\begin{itemize}
%	\item mtrue
%	\item utrue (no arrows)
%	\item utrue (yes arrows)
%	\item reconstruction iterations 1,3,4,8: PCH vs REG
%	\item urecovered (arrows, 3D)
%	\item morozov plot
%	\item Newton convergence (iter, numcg, numPDEsolves (num preconditioner build), gradnorm)
%	\item impulse response batches
%	\item 
%\end{itemize}

We use our method to approximate the data misfit Gauss-Newton Hessian in an inverse problem governed by a PDE which models steady state ice sheet flow \cite{PetraEtAl12}. Briefly, a sheet of ice is flowing down a mountain (see Figure \ref{fig:ice_mountain_mesh}). Given observations of the tangential component of the ice velocity on the top surface of the ice, we invert for the logarithm of the unknown spatially varying basal friction parameter field, which governs the resistance to sliding along the base of the ice sheet. The inverse problem considered here is synthetic, but in practice surface velocity observations are available from satellites.

@@

\begin{subequations}\label{Stokeseqn}
	\begin{align}
	-\nabla \cdot \stress
	&=\bodyforce\,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:1}\\
	\nabla\cdot \velocity
	&=0\,\,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:2}\\
	\stress\normalvec
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{t}, \label{Stokeseqn:3}\\
	\velocity \cdot\normalvec =0 \quad\text{and}\quad \tangentop\left(\stress\normalvec+\exp\left(\basalfriction\right)\velocity\right)
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{b},\label{Stokeseqn:4} \\
	\stress\normalvec+\stokesrobincoeff \velocity &=0\,\,\,\,\,\,\text{ on }\Gamma_{l}.\label{Stokeseqn:5}	
	\end{align}
\end{subequations}
The solution to these equations is the pair $(\velocity, \pressure)$, where $\velocity$ is the ice flow velocity field, and $\pressure$ is the pressure field. Here, $\basalfriction$ is the unknown logarithmic
basal friction field (note that large $\basalfriction$ corresponds to high resistance to sliding) which is defined on the $2D$ surface $\Gamma_b$. The quantity $\bodyforce$ is the body force density due to gravity, $\stokesrobincoeff=10^6$ is a Robin boundary condition constant,  $\normalvec$ is the outward normal and $\tangentop$ is the tangential projection operator that restricts a vector field to its tangential component along the boundary. 
We employ Glen's flow law \cite{Glen1955}, $\stress= 2\eta
\dot{\strain} -I\pressure$,
which is a constitutive law for ice that relates the stress
tensor, $\stress$, to the strain rate tensor,
$\dot{\strain}= \frac{1}{2}\left(
\nabla\velocity+\nabla\velocity^{\top} \right)$. Here $\eta$ is the effective viscosity and $I$ is the unit matrix. Glen's exponent has been chosen to be one, which provides for a linear Stokes model. Note that while the PDE is linear, the parameter-to-solution map, $\basalfriction \mapsto (\velocity, \pressure)$, is nonlinear.

The pressure, $\pressure$, is discretized with first order scalar continuous Galerkin finite elements defined on a mesh of tetrahedra. The velocity, $\velocity$, is discretized with second order continuous Galerkin finite elements on the same mesh. The parameter $\basalfriction$ is discretized with first order scalar continuous Galerkin finite elements on the mesh of triangles that results from restricting the tetrahedral mesh to the basal boundary, $\Gamma_b$. Note that $\Gamma_b$ is a 2D surface embedded in 3D due to the mountain topography. We also generate a flattened version of $\Gamma_b$, denoted by $\Omega \subset \mathbb{R}^2$, by ignoring the height coordinate. The parameter $\basalfriction$ is dually viewed as a function on $\Gamma_b$ for the purpose of solving the Stokes equations, and as a function on $\Omega$ for the purpose of building our Hessian approximations and defining the regularization. 
%This flattening is necessary because our operator approximation method requires adding points on this surface together to get other points on the surface, and this is only well-defined in flat space.

@@

The observations are generated by adding $1\%$ multiplicative
Gaussian noise to the tangential component of the velocity field
restricted to the top surface of the geometry. The true basal sliding
coefficient field and the true velocity field , which is obtained by
solving~\eqref{Stokeseqn}, are shown in Figure~\ref{fig:true_beta_u}.

To reconstruct the basal sliding parameter field $\basalfriction$, we
formulate an inverse problem governed by the Stokes
problem~\eqref{Stokeseqn}. In particular this problem is formulated as
a nonlinear least squares optimization problem, whose cost function
consists of a misfit term (between the observations and model output)
and a bi-Laplacian regularization term
following~\cite{IsaacPetraStadlerEtAl15}. The regularization is
centered at the constant function $q_0(x) = 10.5$. To mitigate
boundary effects we use a constant coefficient Robin boundary
condition as in~\cite{Roininen14}. The parameters are chosen so that
the Green's function of the Hessian of the regularization has a
characteristic length of $0.25$ of the radius of the domain. For the
specific setup, we refer the reader
to~[sec.22]\cite{VillaPetraGhattas21}. We choose the regularization
parameter, $\gamma$, that controls the overall strength of the
regularization using Morozov's discrepancy principle. The optimal
value of $\gamma$ is $7.3 \times 10^3$, which is used for all the
numerical results except for Figure~\ref{fig:gamma_sweep}, where we
vary $\gamma$ to study how changing $\gamma$ impacts the effectiveness
of the proposed preconditioner.

@@

We solve the inverse problem with an inexact Newton preconditioned
conjugate gradient (PCG) scheme and Armijo line search for
globalization~\cite{NocedalWright99}. The (discrete) Newton updates,
$\bm{\searchdir}_k$, are obtained by solving 
\begin{equation}
	\label{eq:newton_system}
	\bm{H} \bm{\searchdir}_k = - \bm{g}_k \qquad \text{or} \qquad \bm{H}_\text{gn} \bm{\searchdir}_k = - \bm{g}_k,
\end{equation}
The initial guess is chosen as the discretization of the constant
function $q_0$. Here $\bm{g}_k$, $\bm{H}$ and $\bm{H}_\text{gn}$ are
the discretized gradient Hessian and Gauss-Newton Hessian of the
inverse problem cost function, respectively.
%The Gauss-Newton Hessian is Hessian of $J$, except in the data misfit, $J_d$, the nonlinear mapping $\basalfriction \mapsto \tangentop \velocity$ is replaced by its local linearization at $\basalfriction_k$.
To ensure positive definiteness of the Hessian we use
$\bm{H}_\text{gn}$ for the first five iterations, and $\bm{H}$ for all
subsequent iterations. The Newton iterations are terminated when
$||\bm{g}_k||/||\bm{g}_0|| < 10^{-8}$. Systems
\eqref{eq:newton_system} are solved inexactly using an inner PCG
iteration, which terminates when the norm of the residual for the
linear system is less than $\min \left(0.5,
\sqrt{||\bm{g}_k||}\right)||\bm{g}_k||$. %PCG only requires applying
%$\bm{H}$ (or $\bm{H}_\text{gn}$) to vectors. Each application of
%$\bm{H}$ (or $\bm{H}_\text{gn}$) to a vector requires solving one
%incremental forward and one incremental adjoint Stokes PDE of the form
%\eqref{Stokeseqn}, but with different right hand sides.
\begin{figure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m1_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter1}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m3_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter2}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m4_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter3}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m6_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter4}
	\end{subfigure}
	%\begin{subfigure}{0.05\textwidth}
	%	\includegraphics[scale=0.1]{mColorBar.png}
	%\end{subfigure}
	\caption{The log basal friction parameter after $1$ (left), $3$ (second from left), and $4$ (third from left) Newton iterations, and the final Newton iterate (right). The PSF (5) preconditioner is constructed between Newton iterations 3 and 4.}
	\label{fig:newton_iterates} 
\end{figure}

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit) using 1, 5, and 25 batches. These approximations are denoted by PSF (1), PSF (5), and PSF (25), respectively. The resulting H-matrices, which may be indefinite due to approximation error, are modified to be symmetric positive definite via the procedure described in Appendix \ref{sec:make_spd}, with $l=2$. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel, especially when one is away from the optimal point. Our results show that our preconditioners are also good preconditioners for $\bm{H}$. The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to an H-matrix approximation of the Hessian of the regularization term to form an H-matrix approximation of $\bm{H}_\text{gn}$.
%
%Straightforward calculation shows that the discretized Hessian of the regularization term is given by
%$
%(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
%$
%where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques.
%
We note that the Hessian of the regularization is a combination of
stiffness and mass matrices, which are sparse. Therefore, we form
H-matrix representations of these matrices and combine them using
standard sparse H-matrix techniques.  We factor the overall
Gauss-Newton Hessian approximation, denoted $\bm{\preconditioner}$,
using fast H-matrix methods, then use the factorization as a
preconditioner.

>>>>

Recently there have been improvements in matrix-free H-matrix construction methods in which an operator is applied to structured random vectors, and the response of the operator to those random vectors is processed to construct an H-matrix approximation  \cite{LevittMartinsson22,LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}. 
These methods (which we do not use here) have been used to approximate Hessians in PDE constrained inverse problems \cite{AmbartsumyanEtAl20,HartlandEtAl23}. 
Although the required number of matrix-vector products grows nearly linearly in the discretized parameter dimension and linearly in the rank of the off-diagonal blocks, the required number of matrix-vector products is still large (e.g., hundreds to thousands).
%These methods are asymptotically scalable in the sense that the required number of matrix-vector products grows nearly linearly in the dimension of the discretized parameter and linearly in the rank of the off-diagonal blocks. But despite asymptotic scalability, from a practical perspective the required number of matrix-vector products is large (e.g., hundreds to thousands). 
%Improving the peeling process to reduce its cost is currently an area of active research \cite{PEELINGRESEARCH}.
%This is an area of active research, so more efficient and practical matrix-free H-matrix construction methods based on the peeling process may be discovered in the future.
%Matrix-free H-matrix construction methods are currently an area of active research (see, e.g., \cite{LevittMartinsson22}), so more efficient matrix-free H-matrix methods may be discovered in the future. 
In this paper, we also form an H-matrix Hessian approximation. However, to reduce the required number of operator actions, we first form a PSF approximation of the data misfit Hessian by exploiting locality and non-negative integral kernel properties, then form the H-matrix using classical H-matrix techniques. Not all data misfit Hessians satisfy the local non-negative integral kernel properties (for example, the wave inverse problem data misfit Hessian has a substantial amount of negative entries in its integral kernel). However, many data misfit Hessians of practical interest do satisfy these properties (either exactly or approximately), and our method is targeted at approximating these Hessians.

@@

%\textbf{Heat equation}
%\begin{itemize}
%	\item Hessian per-column error plots for 1, 5, and 15 batches
%	\item $H-P$ error vs number of batches for interior and whole domain
%	\item $P^{-1}H-I$: error vs rank for $P=R$ and $P=P_\text{pch}$ diffusion time, for 1, 5, and 15 batches
%	\item Krylov iterations to tols 1e-1 and 1e-6 vs diffusion parameter
%	\item GLR vs PCH vs diffusion time
%	\item Krylov method convergence plot: $R$ vs $P_\text{pch}$ vs no preconditioner
%	\item Krylov iterations to tols 1e-6 vs mesh size $h$, PCH vs R vs no preconditioner (hold)
%	\item true parameter (H)
%	\item deterministic reconstruction (H)
%	\item noisy measurements (1\%, 5\%, 10\% noise). $u|_\text{top}$ (H)
%	\item recovered state at top (H)
%	\item mesh scalability of PCH (CC) (H)
%	\item put plot data into data directory
%	\item save function .pvd files in paraview  directory
%\end{itemize}
%
%\begin{itemize}
%	\item GLR vs PCH number of obs (Computational cost in PDE solves) (S)
%	\item GLR vs PCH aspect ratio (CC) (S)
%	\item turn on preconditioner after number of krylov iterations exceeds 15 in an iteration. Rebuild "as needed". (S)
%\end{itemize}
%
%\begin{itemize}
%	\item Stokes Krylov convergence (10 solves for nonlinear forward + 6 hessian matvecs for ellipsoid estimation + 10 matvecs for impulse responses + 30 matvecs for PCH6 CG solve + 100 matvecs for Reg CG solve)
%	\item Stokes error vs num batches (small number of batches; 1, 3, 6, 9 batches?) (50 matvecs for randomized error estimation +)
%	\item Velocity field
%	\item basal sliding field
%	\item Picture of impulse responses
%	\item Mesh negative numbers picture
%\end{itemize}
%
%\begin{itemize}
%	\item mtrue
%	\item utrue (no arrows)
%	\item utrue (yes arrows)
%	\item reconstruction iterations 1,3,4,8: PCH vs REG
%	\item urecovered (arrows, 3D)
%	\item morozov plot
%	\item Newton convergence (iter, numcg, numPDEsolves (num preconditioner build), gradnorm)
%	\item impulse response batches
%	\item 
%\end{itemize}

We use our method to approximate the data misfit Gauss-Newton Hessian in an inverse problem governed by a PDE which models steady state ice sheet flow \cite{PetraEtAl12}. Briefly, a sheet of ice is flowing down a mountain (see Figure \ref{fig:ice_mountain_mesh}). Given observations of the tangential component of the ice velocity on the top surface of the ice, we invert for the logarithm of the unknown spatially varying basal sliding parameter field, which governs the resistance to sliding along the base of the ice sheet. The inverse problem considered here is synthetic, but in practice surface velocity observations are available from satellites.

@@

\begin{subequations}\label{Stokeseqn}
	\begin{align}
	-\nabla \cdot \stress
	&=\bodyforce\,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:1}\\
	\nabla\cdot \velocity
	&=0\,\,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:2}\\
	\stress\normalvec
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{t}, \label{Stokeseqn:3}\\
	\velocity \cdot\normalvec =0 \quad\text{and}\quad \tangentop\left(\stress\normalvec+\exp\left(\basalfriction\right)\velocity\right)
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{b},\label{Stokeseqn:4} \\
	\stress\normalvec+\stokesrobincoeff \velocity &=0\,\,\,\,\,\,\text{ on }\Gamma_{l}.\label{Stokeseqn:5}	
	\end{align}
\end{subequations}
The solution to these equations is the pair $(\velocity, \pressure)$, where $\velocity$ is the ice flow velocity field, and $\pressure$ is the pressure field. Here, $\basalfriction$ is the unknown logarithmic
basal sliding field (note that large $\basalfriction$ corresponds to high resistance to sliding) which is defined on the $2$D surface $\Gamma_b$. The quantity $\bodyforce$ is the body force density due to gravity, $\stokesrobincoeff=10^6$ is a Robin boundary condition constant,  $\normalvec$ is the outward normal and $\tangentop$ is the tangential projection operator that restricts a vector field to its tangential component along the boundary. 
We employ Glen's flow law \cite{Glen1955}, $\stress= 2\eta
\dot{\strain} -I\pressure$,
which is a constitutive law for ice that relates the stress
tensor, $\stress$, to the strain rate tensor,
$\dot{\strain}= \frac{1}{2}\left(
\nabla\velocity+\nabla\velocity^{\top} \right)$. Here $\eta$ is the effective viscosity and $I$ is the unit matrix. Glen's exponent has been chosen to be one, which provides for a linear Stokes model. Note that while the PDE is linear, the parameter-to-solution map, $\basalfriction \mapsto (\velocity, \pressure)$, is nonlinear.

The pressure, $\pressure$, is discretized with first order scalar continuous Galerkin finite elements defined on a mesh of tetrahedra. The velocity, $\velocity$, is discretized with second order continuous Galerkin finite elements on the same mesh. The parameter $\basalfriction$ is discretized with first order scalar continuous Galerkin finite elements on the mesh of triangles that results from restricting the tetrahedral mesh to the basal boundary, $\Gamma_b$. Note that $\Gamma_b$ is a 2D surface embedded in 3D due to the mountain topography. We also generate a flattened version of $\Gamma_b$, denoted by $\Omega \subset \mathbb{R}^2$, by ignoring the height coordinate. The parameter $\basalfriction$ is dually viewed as a function on $\Gamma_b$ for the purpose of solving the Stokes equations, and as a function on $\Omega$ for the purpose of building our Hessian approximations and defining the regularization. 
%This flattening is necessary because our operator approximation method requires adding points on this surface together to get other points on the surface, and this is only well-defined in flat space. 
The observations are generated by adding $1\%$ multiplicative
Gaussian noise to the tangential component of the velocity field
restricted to the top surface of the geometry. The true basal sliding
coefficient field and the true velocity field, which is obtained by
solving~\eqref{Stokeseqn}, are shown in Figure~\ref{fig:true_beta_u}.

@@

To reconstruct the basal sliding parameter field $\basalfriction$, we
formulate an inverse problem governed by the Stokes
problem~\eqref{Stokeseqn}. In particular this problem is formulated as
a nonlinear least squares optimization problem, whose cost function
consists of a misfit term (between the observations and model output)
and a bi-Laplacian regularization term
following~\cite{IsaacPetraStadlerEtAl15}. The regularization is
centered at the constant function $q_0(x) = 10.5$. To mitigate
boundary effects we use a constant coefficient Robin boundary
condition as in~\cite{Roininen14}. The parameters are chosen so that
the Green's function of the Hessian of the regularization has a
characteristic length of $0.25$ of the domain radius. For the
specific setup, we refer the reader
to~\cite[Section 2.2]{VillaPetraGhattas21}. We choose the regularization
parameter, $\gamma$, that controls the overall strength of the
regularization using Morozov's discrepancy principle. The optimal
value of $\gamma$ is $7.3 \times 10^3$, which is used for all the
numerical results except for Figure~\ref{fig:gamma_sweep}, where we
vary $\gamma$ to study how changing $\gamma$ impacts the effectiveness
of the proposed preconditioner.

@@

We solve the inverse problem with an inexact Newton preconditioned
conjugate gradient (PCG) scheme and a globalizing Armijo line search~\cite{NocedalWright99}. The (discrete) Newton updates,
$\bm{\searchdir}_k$, are obtained by solving 
%
\begin{equation}
	\label{eq:newton_system}
	\bm{H} \bm{\searchdir}_k = - \bm{g}_k \qquad \text{or} \qquad \bm{H}_\text{gn} \bm{\searchdir}_k = - \bm{g}_k,
\end{equation}
%
wherein we choose the initial guess as the discretization of the constant
function $q_0$. Here $\bm{g}_k$, $\bm{H}$ and $\bm{H}_\text{gn}$ are
the discretized gradient, Hessian, and Gauss-Newton Hessian of the
inverse problem cost function, respectively.
%The Gauss-Newton Hessian is Hessian of $J$, except in the data misfit, $J_d$, the nonlinear mapping $\basalfriction \mapsto \tangentop \velocity$ is replaced by its local linearization at $\basalfriction_k$.
To ensure positive definiteness of the Hessian we use
$\bm{H}_\text{gn}$ for the first five iterations, and $\bm{H}$ for all
subsequent iterations. The Newton iterations are terminated when
$||\bm{g}_k||/||\bm{g}_0|| < 10^{-8}$. Systems~\eqref{eq:newton_system} are solved inexactly using an inner PCG
iteration, which terminates when the norm of the linear system residual is less than $\min \left(0.5,
\sqrt{||\bm{g}_k||}\right)||\bm{g}_k||$. %PCG only requires applying
%$\bm{H}$ (or $\bm{H}_\text{gn}$) to vectors. Each application of
%$\bm{H}$ (or $\bm{H}_\text{gn}$) to a vector requires solving one
%incremental forward and one incremental adjoint Stokes PDE of the form
%\eqref{Stokeseqn}, but with different right hand sides.
\begin{figure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m1_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter1}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m3_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter2}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m4_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter3}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m6_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter4}
	\end{subfigure}
	%\begin{subfigure}{0.05\textwidth}
	%	\includegraphics[scale=0.1]{mColorBar.png}
	%\end{subfigure}
	\caption{The log basal sliding parameter after $1$ (left), $3$ (second from left), and $4$ (third from left) Newton iterations, and the final Newton iterate (right). The PSF (5) preconditioner is constructed between Newton iterations 3 and 4.}
	\label{fig:newton_iterates} 
\end{figure}

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit) using 1, 5, and 25 batches. These approximations are denoted by PSF (1), PSF (5), and PSF (25), respectively. The resulting H-matrices, which may be indefinite due to approximation error, are modified to be symmetric positive definite via the procedure described in Appendix~\ref{sec:make_spd}, with $l=2$. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel, especially when one is away from the optimal basal sliding parameter. Our results show that our preconditioners are also good preconditioners for $\bm{H}$. The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to an H-matrix approximation of the Hessian of the regularization term to form an H-matrix approximation of $\bm{H}_\text{gn}$.
%
%Straightforward calculation shows that the discretized Hessian of the regularization term is given by
%$
%(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
%$
%where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques.
%
We note that the Hessian of the regularization is a combination of
stiffness and mass matrices, which are sparse. Therefore, we form
H-matrix representations of these matrices and combine them using
standard sparse H-matrix techniques.  We factor the overall
Gauss-Newton Hessian approximation, denoted $\bm{\preconditioner}$,
using fast H-matrix methods, then use the factorization as a
preconditioner.
